---
title: "example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  eval = FALSE, #this makes so that chunks are not rerun 
  cache = TRUE,
  warning = FALSE)
```

# Introduction

# Installation

## Python requirements 
The feature selection procedure is implemented with Python's scikit-learn library. The package requires an Anaconda installation to work properly. The Python requirements are listed in the requirements.txt file. 

You can install the development version of GeneSelectR from [GitHub](%5Bhttps://github.com/)](<https://github.com/>)) with:

NB! Prior to installation you should run this:

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ViSEAGO")
```

And then you can install the development version of the package from here:

```{r}
# install.packages("devtools")
devtools::install_github("dzhakparov/GeneSelectR", build_vignettes = FALSE)

```

# Before you begin 
GeneSelectR is based on usage of Python for feature selection via reticulate. When you install the library and launch it for the first time, it is necessary to create the working environment for the package. 
To do this there is a convenience function: 
```{r}
GeneSelectR::configure_environment() 
```
By running it you will be prompted to install a new conda environment with the default name 'GeneSelectR_env' that will be used in your future GeneSelectR sessions. After that you will be asked if you want necessary Python packages to be installed in this working environment. 

Due to restrictions the reticulate package has after this initial step is done, it is necessary to start every of your GeneSelectR session with setting up the correct conda working environment: 

```{r}
GeneSelectR::set_reticulate_python()
library(GeneSelectR)
# rest of your code 
```
Additionally, number arguments in fit_and_evaluate_pipelines function should be passed as literal integers with an L suffix, e.g: 

```{r}
fit_and_evaluate_pipelines(X_train, 
                           y_train,
                           njobs = 3L) # number of cores to be used for the feature selection process 
```


# Basic Example
## Input Data Requirements
The data matrix should be a data frame with samples as rows and features as columns. Example dataset can be accessed via: 
```{r}
data("UrbanRandomSubset")
head(UrbanRandomSubset[,1:10])
```
This dataset is a bulk RNAseq dataset that was obtained from blood samples of the 149 African children that were stratified into ones having Atopic Dermatitis (AD) and healthy controls (HC). Additionally, the whole dataset contains the stratification variable by the childrens' location (Urban and Rural). This data snippet contains the Urban samples only. The columns represent the genes and samples are in rows. The column treatment contains diangosis label.

This data subset was used in the PharML 2022 workshop within ECML-PKDD 2022 conference. The whole paper is available with the doi: 

## Feature Selection Procedure Basic Usage
As noted before every new session in which you use GeneSelectR should start with the following line, that would allow you to set the correct conda environment:

```{r}
GeneSelectR::set_reticulate_python()
library(GeneSelectR)
library(dplyr)
# the rest of the code
```
Prior to running the feature selection function, let's prepare the data to have a suitable format: 
```{r}
X <- UrbanRandomSubset %>% dplyr::select(-treatment) # get the feature matrix
y <- UrbanRandomSubset['treatment'] # store the data point label in a separate vector
```

To run the feature selection procedure, you should run the fit_and_evaluate_pipelines() function: 
```{r}
selection_results <- fit_and_evaluate_pipelines(X_train = X, 
                                                y_train = y,
                                                njobs = 1L)
selection_results
```
This will run a feature selection procedure for the specified dataset. By default there are three methods implemented to select the features: Univariate (sklearn ref), Logistic regression with L1 penalty (sklearn-ref) and boruta (link to the github page). 
If you wish to add any other methods from sklearn, you should pass it as a named list like here: 
```{r}
# import the sklearn module
sklearn <- reticulate::import('sklearn')
# define the feature selection submodule and wanted methods with an estimator
feature_selection <- sklearn$feature_selection
select_from_model <- feature_selection$SelectFromModel
RFE <- feature_selection$RFE
rf <- sklearn$ensemble$RandomForestClassifier

my_methods <- list('RFE' = RFE(estimator = rf(), n_features_to_select = 100L),
                   'SelectFromModel' = select_from_model(estimator = rf()))
```

And then we can pass it as an argument to fit_and_evaluate_pipelines function: 
```{r}
selection_results <- fit_and_evaluate_pipelines(X_train = X, 
                                                y_train = y,
                                                njobs = -1L,
                                                feature_selection_methods = my_methods)
```
After that the feature selection is performed in a following manner: 
(Block scheme here from pipeline html report with sklearn)
You can specify your estimator of choice as well as parameter grid. 
For the sake of computational time the RandomizedSearchCV is used. 

```{r eval=TRUE}
selection_results
```

The selection_results object belongs to the PipelineResults class and contains the following fields: 
1. Fitted Pipelines 
2. cv_results
3. selected_features 
4. mean_performance (to be removed)
5. mean_feature_importances
6. Gene set stability score (to be added)

## ML Performance Metrics
## Overlap Between Gene Lists

## Go Enrichment analysis
## Conversion to ENTREZIDs
