---
title: "example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  #eval = FALSE this makes so that chunks are not rerun 
  cache = TRUE,
  warning = FALSE)
```

```{r setup}
library(GeneSelectR)
```

# Introduction

# Installation

## Python requirements 
The feature selection procedure 

You can install the development version of GeneSelectR from [GitHub](%5Bhttps://github.com/)](<https://github.com/>)) with:

NB! Prior to installation you should run this:

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ViSEAGO")
```

And then you can install the development version of the package from here:

```{r}
# install.packages("devtools")
devtools::install_github("dzhakparov/GeneSelectR")

```



# Basic Example

## Input Data Requirements
The data matrix should be a data frame with samples as rows and features as columns. Example dataset can be accessed via: 
```{r}
data("UrbanRandomSubset")
head(exp_labeled[,1:10])
```


## Feature Selection Procedure
By calling the package, it will create the conda environment and install all the necessary Python packages for it. Conda installation should be present prior to the installation of the package. 

```{r}
library(GeneSelectR)
library(dplyr)
```
Prior to running the feature selection function, let's prepare the data to have a suitable format: 
```{r}
X <- exp_labeled %>% dplyr::select(-treatment) # get the feature matrix
y <- exp_labeled['treatment'] # store the data point label in a separate vector
```

To run the feature selection procedure, you should run the fit_and_evaluate_pipelines() function: 
```{r}
selection_results <- fit_and_evaluate_pipelines(X_train = X, 
                                                y_train = y,
                                                njobs = 2L)
selection_results
```


## Overlap Between Gene Lists

## Conversion to ENTREZIDs
