---
title: "GeneSelectR 2.0: Gene Selection for Immunotherapy Response in Bladder Cancer"
author: "Damir Zhakparov"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{GeneSelectR 2.0: Gene Selection for Immunotherapy Response in Bladder Cancer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

GeneSelectR 2.0 identifies biologically relevant gene signatures by integrating
three complementary pillars:

- **Selection stability** ($\pi$): How consistently a gene is selected across
  repeated cross-validation folds using regularised regression.
- **Predictive utility** ($u$): A blend of regularised coefficient magnitude
  and mutual information with the outcome.
- **Biological relevance** ($b$): Gene Ontology semantic similarity to
  user-specified or data-driven target processes.

This vignette demonstrates the full workflow on the **IMvigor210** bladder
cancer cohort from the `easierData` Bioconductor package — a dataset of 192
patients treated with anti-PD-L1 (atezolizumab) immunotherapy, with RNA-seq
expression and clinical response annotations.

## Installation

```{r install, eval = FALSE}
# Install Bioconductor dependencies (one-time)
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "easierData",
  "SummarizedExperiment",
  "org.Hs.eg.db",
  "GO.db",
  "AnnotationDbi"
))

# Install GeneSelectR (from GitHub or local)
# devtools::install_github("your-org/GeneSelectR")
```

## Load the IMvigor210 Dataset

The `easierData` package provides the Mariathasan et al. (2018) dataset as a
`SummarizedExperiment` with 31,087 genes across 192 patients. The `colData`
includes the Best Overall Response (BOR) to atezolizumab treatment.

```{r load-data}
library(SummarizedExperiment)
library(easierData)
library(GeneSelectR)

# Load the IMvigor210 bladder cancer dataset
se <- easierData::get_Mariathasan2018_PDL1_treatment()
se
```

```{r inspect-coldata}
# Inspect clinical annotations
table(se$BOR)
```

The BOR column encodes response categories: R - responders, NR - non-responders and NE (not
evaluable).

## Preprocessing

GeneSelectR expects a numeric matrix `X` (samples x genes) and a binary factor
`y`. We need to:

1. Create a binary response label (responder vs non-responder).
2. Filter to protein-coding genes with sufficient expression and variance.
3. Extract a log-TPM matrix.

```{r preprocess}
# --- Step 1: Binary response ---
# Responders: CR + PR; Non-responders: SD + PD
# Remove NE (not evaluable) patients
response_map <- c(R = "Responder", NR = "NonResponder")

keep_patients <- se$BOR %in% names(response_map)
se <- se[, keep_patients]
response <- factor(response_map[se$BOR], levels = c("NonResponder", "Responder"))
table(response)
```

```{r filter-genes}
# --- Step 2: Gene filtering ---
tpm <- assay(se, "tpm")

# Keep genes expressed in at least 10% of samples (TPM > 1)
expressed <- rowSums(tpm > 1) >= ncol(tpm) * 0.10
tpm_filt <- tpm[expressed, ]

# Remove low-variance genes (bottom 50% by variance)
gene_vars <- apply(log1p(tpm_filt), 1, var)
high_var <- gene_vars >= median(gene_vars)
tpm_filt <- tpm_filt[high_var, ]

cat(sprintf("Genes after filtering: %d (from %d)\n", nrow(tpm_filt), nrow(tpm)))
```

```{r prepare-matrix}
# --- Step 3: Log-TPM matrix (samples x genes) ---
X <- t(log1p(tpm_filt))   # samples in rows, genes in columns
y <- response

cat(sprintf("X dimensions: %d samples x %d genes\n", nrow(X), ncol(X)))
cat(sprintf("Class balance: %s\n",
            paste(names(table(y)), table(y), sep = "=", collapse = ", ")))
```

## Run GeneSelectR 2.0

### Baseline: No biological scoring

First, we run GeneSelectR using only the statistical pillars ($\pi$ and $u$),
without GO-based biology scoring. This serves as a baseline.

```{r fit-none, cache = TRUE}
result_none <- geneselectr2_fit(
  X = X,
  y = y,
  bio_mode = "none",
  regularization_method = "elastic_net",
  alpha = 0.5,
  K = 5,
  R = 10,
  random_seed = 42,
  verbose = TRUE
)
```

```{r results-none}
# Cross-validation performance
cat(sprintf("Mean AUC: %.4f +/- %.4f\n",
            result_none$cv_summary$mean,
            result_none$cv_summary$sd))

# Top 20 genes (ranked by pi x u)
head(result_none$gene_scores[, c("gene", "final_score", "pi_exact", "u")], 20)
```

### Supervised mode: immune-related GO terms

In supervised mode, we provide specific GO terms that encode our biological
hypothesis — here, that response to immunotherapy is driven by immune processes.

```{r fit-supervised, cache = TRUE}
# Key immune-related GO terms
immune_terms <- c(
  "GO:0006955",  # immune response
  "GO:0006954",  # inflammatory response
  "GO:0042110",  # T cell activation
  "GO:0002250",  # adaptive immune response
  "GO:0019882",  # antigen processing and presentation
  "GO:0002407"   # dendritic cell chemotaxis
)

result_sup <- geneselectr2_fit(
  X = X,
  y = y,
  bio_mode = "supervised",
  target_terms = immune_terms,
  bio_ontology = "BP",
  bio_sim_method = "resnik",
  regularization_method = "elastic_net",
  alpha = 0.5,
  K = 5,
  R = 10,
  random_seed = 42,
  verbose = TRUE
)
```

```{r results-supervised}
head(result_sup$gene_scores[, c("gene", "final_score", "pi_exact", "u", "b")], 20)
```

### Data-driven mode: auto-detect relevant biology

In data-driven mode, GeneSelectR first identifies a candidate gene set from the
CV results, runs GO enrichment on it, and uses the enriched terms as targets
for biological scoring. This is useful when you don't have a strong prior
hypothesis.

```{r fit-datadriven, cache = TRUE}
result_dd <- geneselectr2_fit(
  X = X,
  y = y,
  bio_mode = "data_driven",
  bio_ontology = "BP",
  bio_sim_method = "lin",
  bio_enrich_fdr = 0.05,
  bio_ic_quantile = 0.5,
  bio_max_enriched = 100,
  bio_n_top_sims = 5,
  regularization_method = "elastic_net",
  alpha = 0.5,
  K = 5,
  R = 10,
  random_seed = 42,
  verbose = TRUE
)
```

```{r results-datadriven}
head(result_dd$gene_scores[, c("gene", "final_score", "pi_exact", "u", "b")], 20)
```

## Comparing Results Across Modes

```{r compare-modes}
# Extract top-50 gene sets from each mode
top_n <- 50
top_none <- head(result_none$gene_scores$gene, top_n)
top_sup  <- head(result_sup$gene_scores$gene, top_n)
top_dd   <- head(result_dd$gene_scores$gene, top_n)

# Overlap summary
cat("=== Top-50 gene overlap ===\n")
cat(sprintf("none & supervised:  %d\n", length(intersect(top_none, top_sup))))
cat(sprintf("none & data-driven: %d\n", length(intersect(top_none, top_dd))))
cat(sprintf("supervised & data-driven: %d\n", length(intersect(top_sup, top_dd))))
cat(sprintf("All three: %d\n", length(Reduce(intersect, list(top_none, top_sup, top_dd)))))
```

```{r common-genes}
# Genes selected by all three modes
common_all <- Reduce(intersect, list(top_none, top_sup, top_dd))
cat("\nGenes in top 50 across ALL three modes:\n")
print(common_all)
```

## Understanding the Score Components

Each gene receives three normalised scores that are combined into a final
ranking. Let's visualise how these components relate to each other in the
supervised run.

```{r score-components, fig.width = 8, fig.height = 6, echo=TRUE}
gs <- result_sup$gene_scores

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# Stability vs Utility
plot(gs$pi_exact, gs$u, pch = 16, cex = 0.4, col = "#00000040",
     xlab = expression(pi ~ "(selection stability)"),
     ylab = "u (predictive utility)",
     main = "Stability vs Utility")

# Stability vs Biology
plot(gs$pi_exact, gs$b, pch = 16, cex = 0.4, col = "#00000040",
     xlab = expression(pi ~ "(selection stability)"),
     ylab = "b (biological relevance)",
     main = "Stability vs Biology")

# Utility vs Biology
plot(gs$u, gs$b, pch = 16, cex = 0.4, col = "#00000040",
     xlab = "u (predictive utility)",
     ylab = "b (biological relevance)",
     main = "Utility vs Biology")

# Final score distribution
hist(gs$final_score, breaks = 50, col = "steelblue",
     border = "white", main = "Final score distribution",
     xlab = "Final score")
```

## Comparing Score Formulas

GeneSelectR supports three combination formulas. The geometric mean (default)
provides a good balance; the harmonic mean strongly penalises genes with any
weak component; the arithmetic mean is lenient.

```{r compare-formulas, cache = TRUE}
formulas <- c("geometric", "arithmetic", "harmonic")

formula_results <- lapply(formulas, function(f) {
  geneselectr2_fit(
    X = X, y = y,
    bio_mode = "supervised",
    target_terms = immune_terms,
    score_formula = f,
    K = 5, R = 5,
    verbose = FALSE,
    random_seed = 42
  )
})
names(formula_results) <- formulas

# Compare top-20 overlap across formulas
top20_per_formula <- lapply(formula_results, function(r) head(r$gene_scores$gene, 20))

cat("Top-20 overlap matrix:\n")
n_formulas <- length(formulas)
overlap_mat <- matrix(0, n_formulas, n_formulas,
                       dimnames = list(formulas, formulas))
for (i in seq_len(n_formulas)) {
  for (j in seq_len(n_formulas)) {
    overlap_mat[i, j] <- length(intersect(top20_per_formula[[i]],
                                            top20_per_formula[[j]]))
  }
}
print(overlap_mat)
```

## Effect of Custom Weights

You can weight the three pillars differently to emphasise stability,
predictive power, or biological relevance.

```{r custom-weights, cache = TRUE}
# Emphasise biology
result_bio_heavy <- geneselectr2_fit(
  X = X, y = y,
  bio_mode = "supervised",
  target_terms = immune_terms,
  score_weights = c(1, 1, 3),  # triple weight on biology
  K = 5, R = 5,
  verbose = FALSE,
  random_seed = 42
)

# Emphasise stability
result_pi_heavy <- geneselectr2_fit(
  X = X, y = y,
  bio_mode = "supervised",
  target_terms = immune_terms,
  score_weights = c(3, 1, 1),  # triple weight on stability
  K = 5, R = 5,
  verbose = FALSE,
  random_seed = 42
)

cat("Top 10 (biology-heavy):\n")
print(head(result_bio_heavy$gene_scores[, c("gene", "final_score", "pi_exact", "b")], 10))

cat("\nTop 10 (stability-heavy):\n")
print(head(result_pi_heavy$gene_scores[, c("gene", "final_score", "pi_exact", "b")], 10))
```

## Using Lasso and Group Lasso

GeneSelectR supports three regularisation strategies. Group lasso encourages
selection of entire gene groups (e.g., correlated clusters) rather than
individual genes.

```{r regularization, cache = TRUE}
# Standard lasso
result_lasso <- geneselectr2_fit(
  X = X, y = y,
  bio_mode = "none",
  regularization_method = "lasso",
  K = 5, R = 5,
  verbose = FALSE,
  random_seed = 42
)

# Group lasso with auto-generated correlation-based groups
result_grp <- geneselectr2_fit(
  X = X, y = y,
  bio_mode = "none",
  regularization_method = "group_lasso",
  K = 5, R = 5,
  verbose = FALSE,
  random_seed = 42
)

cat(sprintf("Lasso - genes with pi > 0.5: %d\n",
            sum(result_lasso$gene_scores$pi_exact > 0.5)))
cat(sprintf("Group lasso - genes with pi > 0.5: %d\n",
            sum(result_grp$gene_scores$pi_exact > 0.5)))
```

## Working with Biology Parameters

The biology scoring system is highly configurable. Here we demonstrate
key parameters using the supervised mode.

### Ontology selection

By default only Biological Process (BP) is used. You can include Molecular
Function and Cellular Component:

```{r ontology-all, cache = TRUE}
result_all_ont <- geneselectr2_fit(
  X = X, y = y,
  bio_mode = "supervised",
  target_terms = immune_terms,
  bio_ontology = c("BP", "MF", "CC"),
  bio_sim_method = "resnik",
  K = 5, R = 5,
  verbose = FALSE,
  random_seed = 42
)

# Compare biology scores: BP-only vs all ontologies
b_bp  <- result_sup$gene_scores$b
b_all <- result_all_ont$gene_scores$b
cat(sprintf("Correlation between BP-only and all-ontology b scores: %.3f\n",
            cor(b_bp, b_all, use = "complete.obs")))
```

### Similarity metrics

Four semantic similarity metrics are available. Relevance (Schlicker) penalises
shallow common ancestors most aggressively:

```{r sim-methods, cache = TRUE}
methods <- c("resnik", "lin", "jiang", "rel")

sim_results <- lapply(methods, function(m) {
  geneselectr2_fit(
    X = X, y = y,
    bio_mode = "supervised",
    target_terms = immune_terms,
    bio_sim_method = m,
    K = 5, R = 3,
    verbose = FALSE,
    random_seed = 42
  )
})
names(sim_results) <- methods

# Extract biology scores and correlate
b_matrix <- sapply(sim_results, function(r) r$gene_scores$b)
colnames(b_matrix) <- methods
cat("Biology score correlations across similarity metrics:\n")
print(round(cor(b_matrix, use = "complete.obs"), 3))
```

## Cache Management

GeneSelectR caches GO annotations and ancestor maps to disk for fast reuse.

```{r cache-management, eval = FALSE}
# View cache statistics
cache_info()

# Clear all cached data (e.g., after updating GO.db)
clear_cache(confirm = FALSE)
```

## Extracting a Gene Signature

A practical use case is to extract a compact gene signature for downstream
use (e.g., building a classifier or scoring new samples).

```{r extract-signature}
gs <- result_sup$gene_scores

# Strategy 1: genes with high stability (pi > 0.5)
sig_stable <- gs$gene[gs$pi_exact > 0.5]
cat(sprintf("Stable signature (pi > 0.5): %d genes\n", length(sig_stable)))

# Strategy 2: top N by final score
sig_top50 <- head(gs$gene, 50)

# Strategy 3: genes in top quartile of all three components
q75_pi <- quantile(gs$pi_exact, 0.75)
q75_u  <- quantile(gs$u, 0.75)
q75_b  <- quantile(gs$b, 0.75)

sig_triple <- gs$gene[gs$pi_exact >= q75_pi & gs$u >= q75_u & gs$b >= q75_b]
cat(sprintf("Triple-quartile signature: %d genes\n", length(sig_triple)))

# Preview the triple-quartile signature
if (length(sig_triple) > 0) {
  print(gs[gs$gene %in% sig_triple, c("gene", "final_score", "pi_exact", "u", "b")])
}
```

## Session Info

```{r sessioninfo}
sessionInfo()
```
